---
title: "Project 4 Membereship detection with profile HMMs"
date: "March 23, 2023"
author: Team K - Minghang Li
output:
  pdf_document:
    latex_engine: xelatex
    pandoc_args: "--pdf-engine-opt=-shell-escape"
    extra_dependencies: ["graphviz", "multirow", "float", "kpfonts"]
---

```{r, include=FALSE}
library(foreach)
library(doParallel)
```

# Problem 8: Estimating match emission probabilities

The count $E_i(a)$ and insert emission probability $e_i(a)$, $\mathcal{a} \in \mathcal{A} = \{A, C, G, T\}$ are:

```{=latex}
\setlength{\tabcolsep}{10pt} % Default value: 6pt
\renewcommand{\arraystretch}{1.5}
\begin{table}[ht]
\centering
\begin{tabular}{|l|l|l|}
\hline
\textbf{pos}                & \textbf{$E_i(a)$}             & \textbf{$e_i(a)$}                                 \\ \hline
\multirow{4}{*}{\textbf{1}} & $E_1(A) = 4 + 1 = \mathbf{5}$ & $e_1(A) = \frac{4+1}{4+4} = \mathbf{\frac{5}{8}}$ \\ \cline{2-3}
                            & $E_1(C) = 0 + 1 = \mathbf{1}$ & $e_1(C) = \frac{0+1}{4+4} = \mathbf{\frac{1}{8}}$ \\ \cline{2-3}
                            & $E_1(G) = 0 + 1 = \mathbf{1}$ & $e_1(G) = \frac{0+1}{4+4} = \mathbf{\frac{1}{8}}$ \\ \cline{2-3}
                            & $E_1(T) = 0 + 1 = \mathbf{1}$ & $e_1(T) = \frac{0+1}{4+4} = \mathbf{\frac{1}{8}}$ \\ \hline
\multirow{4}{*}{\textbf{2}} & $E_2(A) = 0 + 1 = \mathbf{1}$ & $e_2(A) = \frac{0+1}{4+4} = \mathbf{\frac{1}{8}}$ \\ \cline{2-3}
                            & $E_2(C) = 0 + 1 = \mathbf{1}$ & $e_2(C) = \frac{0+1}{4+4} = \mathbf{\frac{1}{8}}$ \\ \cline{2-3}
                            & $E_2(G) = 4 + 1 = \mathbf{5}$ & $e_2(G) = \frac{4+1}{4+4} = \mathbf{\frac{5}{8}}$ \\ \cline{2-3}
                            & $E_2(T) = 0 + 1 = \mathbf{1}$ & $e_2(T) = \frac{0+1}{4+4} = \mathbf{\frac{1}{8}}$ \\ \hline
\multirow{4}{*}{\textbf{3}} & $E_3(A) = 0 + 1 = \mathbf{1}$ & $e_3(A) = \frac{0+1}{5+4} = \mathbf{\frac{1}{9}}$ \\ \cline{2-3}
                            & $E_3(C) = 5 + 1 = \mathbf{6}$ & $e_3(C) = \frac{5+1}{5+4} = \mathbf{\frac{2}{3}}$ \\ \cline{2-3}
                            & $E_3(G) = 0 + 1 = \mathbf{1}$ & $e_3(G) = \frac{0+1}{5+4} = \mathbf{\frac{1}{9}}$ \\ \cline{2-3}
                            & $E_3(T) = 0 + 1 = \mathbf{1}$ & $e_3(T) = \frac{0+1}{5+4} = \mathbf{\frac{1}{9}}$ \\ \hline
\end{tabular}
\end{table}
```

# Problem 9: Estimating insert emission probabilities

Since the contiguous insert states have the same position in the model, the count $E_i(a)$ and insert emission probability $e_i(a)$, $\mathcal{a} \in \mathcal{A} = \{A, C, G, T\}$ are:

```{=latex}
\setlength{\tabcolsep}{10pt} % Default value: 6pt
\renewcommand{\arraystretch}{1.5}
\begin{table}[ht]
\centering
\begin{tabular}{|l|l|}
\hline
\textbf{$E(a)$}             & \textbf{$e(a)$}                                 \\ \hline
$E(A) = 5 + 1 = \mathbf{6}$ & $e(A) = \frac{5+1}{5+1+4} = \mathbf{\frac{3}{5}}$ \\ \hline
$E(C) = 0 + 1 = \mathbf{1}$ & $e(C) = \frac{0+1}{5+1+4} = \mathbf{\frac{1}{10}}$ \\ \hline
$E(G) = 1 + 1 = \mathbf{2}$ & $e(G) = \frac{1+1}{5+1+4} = \mathbf{\frac{1}{5}}$ \\ \hline
$E(T) = 0 + 1 = \mathbf{1}$ & $e(T) = \frac{0+1}{5+1+4} = \mathbf{\frac{1}{10}}$ \\ \hline
\end{tabular}
\end{table}
\pagebreak
```

# Problem 10: Estimating transition probabilities

The path of each sequence (\textcolor[HTML]{1f77b4}{bat}, \textcolor[HTML]{ff7f0e}{rat}, \textcolor[HTML]{2ca02c}{cat}, \textcolor[HTML]{d62728}{gnat}, \textcolor[HTML]{9467bd}{goat}) from Begin to end is summarized below,

![HMM.pdf](./HMM.pdf)

Nodes that are not visited in any path have uniform probability of transitting to its next states:

| Node | $t_i(k \rightarrow l)$ |
| --- | --- |
| $I_0$ | $\frac{1}{3}$ |
| $I_1$ | $\frac{1}{3}$ |
| $I_3$ | $\frac{1}{2}$ |
| $D_3$ | $\frac{1}{2}$ |

And the transition probabilities of nodes that are visited are listed in the following table

```{=latex}
\pagebreak
\setlength{\tabcolsep}{10pt} % Default value: 6pt
\renewcommand{\arraystretch}{1.5}
\begin{table}[ht]
\centering
\begin{tabular}{|l|l|l|}
\hline
\textbf{pos}                & \textbf{$T_i(a)$}             & \textbf{$e_i(a)$}                                 \\ \hline
\multirow{3}{*}{\textbf{0}} & $T_0(M \rightarrow M) = 4 + 1 = \mathbf{5}$ & $t_0(M \rightarrow M) = \frac{4+1}{5+1+2} = \mathbf{\frac{5}{8}}$ \\ \cline{2-3}
                            & $T_0(M \rightarrow I) = 0 + 1 = \mathbf{1}$ & $t_0(M \rightarrow I) = \frac{0+1}{5+1+2} = \mathbf{\frac{1}{8}}$ \\ \cline{2-3}
                            & $T_0(M \rightarrow D) = 1 + 1 = \mathbf{2}$ & $t_0(M \rightarrow D) = \frac{0+1}{5+1+2} = \mathbf{\frac{1}{8}}$ \\ \hline
\multirow{6}{*}{\textbf{1}} & $T_1(M \rightarrow M) = 3 + 1 = \mathbf{4}$ & $t_1(M \rightarrow M) = \frac{3+1}{4+1+2} = \mathbf{\frac{4}{7}}$ \\ \cline{2-3}
                            & $T_1(M \rightarrow I) = 0 + 1 = \mathbf{1}$ & $t_1(M \rightarrow I) = \frac{0+1}{4+1+2} = \mathbf{\frac{1}{7}}$ \\ \cline{2-3}
                            & $T_1(M \rightarrow D) = 1 + 1 = \mathbf{2}$ & $t_1(M \rightarrow D) = \frac{0+1}{4+1+2} = \mathbf{\frac{1}{7}}$ \\ \cline{2-3}
                            & $T_1(D \rightarrow M) = 1 + 1 = \mathbf{2}$ & $t_1(D \rightarrow M) = \frac{1+1}{2+1+1} = \mathbf{\frac{2}{4}}$ \\ \cline{2-3}
                            & $T_1(D \rightarrow I) = 0 + 1 = \mathbf{1}$ & $t_1(D \rightarrow I) = \frac{0+1}{2+1+1} = \mathbf{\frac{1}{4}}$ \\ \cline{2-3}
                            & $T_1(D \rightarrow D) = 0 + 1 = \mathbf{1}$ & $t_1(D \rightarrow D) = \frac{0+1}{2+1+1} = \mathbf{\frac{1}{4}}$ \\ \hline
\multirow{9}{*}{\textbf{2}} & $T_2(M \rightarrow M) = 2 + 1 = \mathbf{3}$ & $t_2(M \rightarrow M) = \frac{2+1}{3+3+1} = \mathbf{\frac{3}{7}}$ \\ \cline{2-3}
                            & $T_2(M \rightarrow I) = 2 + 1 = \mathbf{3}$ & $t_2(M \rightarrow I) = \frac{2+1}{3+3+1} = \mathbf{\frac{3}{7}}$ \\ \cline{2-3}
                            & $T_2(M \rightarrow D) = 0 + 1 = \mathbf{1}$ & $t_2(M \rightarrow D) = \frac{0+1}{3+3+1} = \mathbf{\frac{1}{7}}$ \\ \cline{2-3}
                            & $T_2(I \rightarrow M) = 3 + 1 = \mathbf{4}$ & $t_2(I \rightarrow M) = \frac{3+1}{4+4+1} = \mathbf{\frac{4}{9}}$ \\ \cline{2-3}
                            & $T_2(I \rightarrow I) = 3 + 1 = \mathbf{4}$ & $t_2(I \rightarrow I) = \frac{3+1}{4+4+1} = \mathbf{\frac{4}{9}}$ \\ \cline{2-3}
                            & $T_2(I \rightarrow D) = 0 + 1 = \mathbf{1}$ & $t_2(I \rightarrow D) = \frac{0+1}{4+4+1} = \mathbf{\frac{1}{9}}$ \\ \cline{2-3}
                            & $T_2(D \rightarrow M) = 0 + 1 = \mathbf{1}$ & $t_2(D \rightarrow M) = \frac{0+1}{1+2+1} = \mathbf{\frac{1}{4}}$ \\ \cline{2-3}
                            & $T_2(D \rightarrow I) = 1 + 1 = \mathbf{2}$ & $t_2(D \rightarrow I) = \frac{1+1}{1+2+1} = \mathbf{\frac{2}{4}}$ \\ \cline{2-3}
                            & $T_2(D \rightarrow D) = 0 + 1 = \mathbf{1}$ & $t_2(D \rightarrow D) = \frac{0+1}{1+2+1} = \mathbf{\frac{1}{4}}$ \\ \hline
\multirow{2}{*}{\textbf{3}} & $T_3(M \rightarrow M) = 5 + 1 = \mathbf{6}$ & $t_3(M \rightarrow M) = \frac{5+1}{6+1} = \mathbf{\frac{6}{7}}$ \\ \cline{2-3}
                            & $T_3(M \rightarrow I) = 0 + 1 = \mathbf{1}$ & $t_3(M \rightarrow I) = \frac{0+1}{6+1} = \mathbf{\frac{1}{7}}$ \\ \hline
\end{tabular}
\end{table}
```

All the transition probabilities are summarized in Figure 2.

![HMM profile](./HMM_profile.pdf)

# Problem 11: Protein family membership classification

## Import functions and read alignments

```{r}
# import functions
source("code/profileHMM.R", local = knitr::knit_global())
```

```{r}
# read alignments
GTPase <- parseAlignment("./data/GTP_binding_proteins.txt")
ATPase <- parseAlignment("./data/ATPases.txt")
```

## Learn HMM from two protein families

```{r}
HMM_GTPase <- learnHMM(GTPase)
HMM_ATPase <- learnHMM(ATPase)
```

## Identify position(s) with the highest match and insert emission frequencies over all symbols.

```{r}
alphabet <- c("A","C","D","E","F","G","H","I","K","L","M", "N","P","Q","R","S","T","V","W","Y")
```

### ATPase

```{r}
mE_max_ATPase <- which(HMM_ATPase$mE == max(HMM_ATPase$mE, na.rm = TRUE), arr.ind = TRUE)
iE_max_ATPase <- which(HMM_ATPase$iE == max(HMM_ATPase$iE, na.rm = TRUE), arr.ind = TRUE)
print(paste("The position(s) with the highest match emission frequency over all symbols is/are:",
            mE_max_ATPase[2]))
print(paste("The position(s) with the highest insert emission frequency over all symbols is/are:",
            iE_max_ATPase[2]))
```

```{r}
# generate barplot color sequence c() with length of alphabet
# all colors are gray except for position at mE_max_ATPase[1] is red
color_mE_max_ATPase <- c(rep("gray", length(alphabet)))
color_mE_max_ATPase[mE_max_ATPase[1]] <- "red"
barplot(HMM_ATPase$mE[, mE_max_ATPase[2]],
        col = color_mE_max_ATPase,
        names.arg = alphabet,
        main = "Match emission frequency distribution at highest mE position",
        xlab = "Amino acids",
        ylab = "Match emission frequency")
```

```{r}
# generate barplot color sequence c() with length of alphabet
# all colors are gray except for position at mE_max_ATPase[1] is red
color_iE_max_ATPase <- c(rep("gray", length(alphabet)))
color_iE_max_ATPase[iE_max_ATPase[1]] <- "red"
barplot(HMM_ATPase$iE[, iE_max_ATPase[2]],
        col = color_iE_max_ATPase,
        names.arg = alphabet,
        main = "Insertion emission frequency distribution at highest iE position",
        xlab = "Amino acids",
        ylab = "Insertion emission frequency")
```
### GTPase

```{r}
mE_max_GTPase <- which(HMM_GTPase$mE == max(HMM_GTPase$mE, na.rm = TRUE), arr.ind = TRUE)
iE_max_GTPase <- which(HMM_GTPase$iE == max(HMM_GTPase$iE, na.rm = TRUE), arr.ind = TRUE)
print(paste("The position(s) with the highest match emission frequency over all symbols is/are:",
            mE_max_GTPase[2]))
print(paste("The position(s) with the highest insert emission frequency over all symbols is/are:",
            iE_max_GTPase[2]))
```


```{r}
# generate barplot color sequence c() with length of alphabet
# all colors are gray except for position at mE_max_ATPase[1] is red
color_mE_max_GTPase <- c(rep("gray", length(alphabet)))
color_mE_max_GTPase[mE_max_GTPase[1]] <- "red"
barplot(HMM_GTPase$mE[, mE_max_GTPase[2]],
        col = color_mE_max_GTPase,
        names.arg = alphabet,
        main = "Match emission frequency distribution at highest mE position",
        xlab = "Amino acids",
        ylab = "Match emission frequency")
```

```{r}
# generate barplot color sequence c() with length of alphabet
# all colors are gray except for position at mE_max_ATPase[1] is red
color_iE_max_GTPase <- c(rep("gray", length(alphabet)))
color_iE_max_GTPase[iE_max_GTPase[1]] <- "red"
barplot(HMM_GTPase$iE[, iE_max_GTPase[2]],
        col = color_iE_max_GTPase,
        names.arg = alphabet,
        main = "Insertion emission frequency distribution at highest iE position",
        xlab = "Amino acids",
        ylab = "Insertion emission frequency")
```

## Classify unclassified proteins

```{r}
unclassified_proteins <- parseProteins(proteinsFile = "./data/Unclassified_proteins.txt")
```

```{r}
num_cores <- detectCores()
registerDoParallel(num_cores)
log_odds_ATPase <- foreach(seq=unclassified_proteins, .combine = c) %dopar% {
  forward(HMM = HMM_ATPase, seq = seq)
}
log_odds_GTPase <- foreach(seq=unclassified_proteins, .combine = c) %dopar% {
  forward(HMM = HMM_GTPase, seq = seq)
}
stopImplicitCluster()
```

```{r}
# plot the log odds ratio of ATPase (red) and GTPase (blue) on the same plot
plot(log_odds_GTPase, col="blue",
     xlab = "Sequence index",
     ylab = "Log odds vs random")
points(log_odds_ATPase, col="red")
# add legend
legend("right", legend=c("ATPase", "GTPase"), col=c("red", "blue"), lty=1, cex=0.8)
```

```{r}
q <- log_odds_ATPase - log_odds_GTPase
plot(q, col=ifelse(q >= 0, "red", "blue"),
     xlab = "Sequence index",
     ylab = "Log odds ATPase vs GTPase")
legend("right", legend=c("ATPase", "GTPase"), col=c("red", "blue"), lty=1, cex=0.8)
```

We can see a clear separation between points that are classified as ATPases and those classified as GTPases. The log ratio of ATPase vs GTPase is always way larger than 0 or way smaller than 0, indicating high likelihood.

```{r}
# print the index of proteins that are classified as ATPases
print("The index of proteins that are classified as ATPases are:")
print(which(q >= 0))
# print the index of proteins that are classified as GTPases
print("The index of proteins that are classified as GTPases are:")
print(which(q < 0))
```
