---
title: "Projecy 7 Exact inference in graphical models"
date:  "`r format(Sys.time(), '%B %d, %Y')`"
author: "Team K - Minghang Li"
output:
  pdf_document:
    latex_engine: xelatex
    extra_dependencies:  [ "pdfpages", "amsthm", "amssymb", "cancel", "kpfonts", "unicode-math" ]
---

```{r, include=FALSE}
library(foreach)
library(doParallel)
```
Problem 17 and 18 are handwritten (not energetic enough for formatting this time
 as I usually do) and are included using `pdfpages`. Thye don't fit under the
title page so I put problem 19 at the beginning.

# Problem 19: Message passing on a chain

## (a) Store clique potentials in an R object

```{r}
# use the hint!
clique_potentials <- array(dim = c(2, 2, 4),
                           dimnames = list(
                             c("0", "1"), c("0", "1"),
                             c("Psi12", "Psi23", "Psi34", "Psi45")))

```

```{r}
X1 <- c(1/3, 2/3)
X2 <- c(4/5, 2/3)
X3 <- c(5/7, 1/3)
X4 <- c(3/5, 2/5)
X5 <- c(1/2, 7/9)
```

```{r}
clique_potentials[ , 1, "Psi12"] <- (1 - X1) * (1 - X2)
clique_potentials[ , 2, "Psi12"] <- (1 - X1) * X2
clique_potentials[ ,  , "Psi23"] <- c(1-X3, X3)
clique_potentials[ ,  , "Psi34"] <- c(1-X4, X4)
clique_potentials[ ,  , "Psi45"] <- c(1-X5, X5)
clique_potentials
```

## (b) Computing forward messages

```{r message=FALSE, results='hide'}
fwd_msg <- array(dim=c(5, 2),
                 dimname=list(
                   c("X1", "X2", "X3", "X4", "X5"),
                   c("0", "1")
                 ))
fwd_msg["X1", ] <- c(1, 1)

foreach(i=2:5) %do% {
  fwd_msg[i, ] <- fwd_msg[i-1, ] %*% clique_potentials[, , i-1]
}
```

```{r}
fwd_msg
```

## (c) Computing backward messages


Initialize in such shape for easy computation :)

```{r message=FALSE, results='hide'}
bkwd_msg <- array(dim=c(2, 5),
                 dimname=list(
                   c("0", "1"),
                   c("X1", "X2", "X3", "X4", "X5")
                 ))
bkwd_msg[, "X5"] <- c(1, 1)
foreach (i=4:1) %do% {
  bkwd_msg[, i] <- clique_potentials[, , i] %*% bkwd_msg[, i+1]
}
```

```{r}
bkwd_msg
```
## (d) Compute the marginal probability distribution for eahc node

```{r results='hide'}
marginal <- array(0, dim=c(5,2),
                  dimname=list(
                    c("P(X1)", "P(X2)", "P(X3)", "P(X4)", "P(X5)"),
                    c("0", "1")))
foreach (i=1:5) %do% {
  marginal[i, ] <- fwd_msg[i, ] * bkwd_msg[, i]
}
Z <- rowSums(marginal)
marginal <- marginal / Z
```
```{r}
marginal
```

```{r echo=FALSE}
print("The normalising constant Z is 1")
print(Z)
```

\includepdf[pages=-]{smcb_project7.pdf}
