---
title: "Project 2 EM Algorithm"
date: "March 24, 2023"
author: Team K
output: pdf_document
---

```{r, include=FALSE}
source("code/viterbi.r", local = knitr::knit_global())
# or sys.source("your-script.R", envir = knitr::knit_global())
```

# Problem 6: Hidden Markov Models

According to slides 18, the HMM is parameterized by:

$$
\begin{aligned}
&\text{Initial state probabilities: }& I_k = P(Z_1 = k) \\
&\text{Transition probabilities: } & T_{kl} = P(Z_n=l|Z_{n-1}=k) \\
&\text{Emission probabilities: } & E_{kx} = P(X_n = x | Z_n = k)
\end{aligned}
$$

## (a)

$I_k$ should sum up to $1$ $\longrightarrow$ degree of freedom = $K-1$

In the $K \times K$ matrix $T$, each row needs to sum up to $1$ $\longrightarrow$ degree of freedom = $K-1$ for each row

Similarly, in the $K \times M$ matrix $X$, each row needs to sum up to $1$ $\longrightarrow$ degree of freedom = $M-1$ for each row

Hence, the maximum number of free parameters is $(K-1) + K \times (K-1) + K \times (M-1)$

## (b)

Computing the stationary distribution is equivalent to solving $\pi^t = \pi^t T$.

Let $\pi^t = \begin{bmatrix}\pi_1 & \pi_2\end{bmatrix})$, we need to solve:

$$
\begin{aligned}
\begin{bmatrix}\pi_1 & \pi_2\end{bmatrix} 
&= \begin{bmatrix}
  \pi_1& \pi_2
\end{bmatrix}\begin{bmatrix}
  0.2 & 0.8 \\
  0.6 & 0.4
\end{bmatrix} \\
&= \begin{bmatrix}
  0.2\pi_1 + 0.6 \pi_2 &
  0.8\pi_1 + 0.4 \pi_2
\end{bmatrix}
\end{aligned}
$$

Given that $\pi_1 = 1 - \pi_2$e can solve the equations:

$$
\left\{\begin{aligned}
  &\pi_1 = 0.2\pi_1 + 0.6\pi_2\\
  &\pi_2 = 0.8\pi_1 + 0.4\pi_2\\
  &\pi_1 = 1 - \pi_2
\end{aligned}\right.
\Rightarrow
\left\{\begin{aligned}
  &\pi_1 = \frac{3}{7}\\
  &\pi_2 = \frac{4}{7}
\end{aligned}\right.
$$

Alternatively we can solve by solving eigens (which is porbably the more proper way to do it because solving the equations above by hand for high dimensional transition matrices is not a very smart idea). Since $\pi^t = \pi^tT$, we can re-write it into $\pi = T^t \pi$, which means $\pi$ is just the corresponding eigenvector of eigenvalue $\lambda = 1$.

Note that the eigenvector should be normalized to satisfy the constraint that $\pi_1 + \pi_2 = 1$.

```{r}
# initialize the transition matrix
T <- matrix(c(0.2, 0.8, 0.6, 0.4), 2, 2, byrow=TRUE)

# solve for the eigenvalues and eigenvectors for transpose of T
eigens <- eigen(t(T))

# get the index of lambda = 1
index <- which(eigens$values == 1)[1]
ev <- eigens$vectors[, index]

# normalize so that the probs sum to 1
pi <- ev / sum(ev)
pi
```

```{r}
# check that the two computation confirms
abs(pi[1] - 3/7) < 1e-5
abs(pi[2] - 4/7) < 1e-5
```

# Problem 7: Predictinig protein secondary structure using HMMs

## (a)

Read `proteins train.tsv`, `proteins test.tsv` and `proteins new.tsv` into the memory and store each in a `data.frame`

```{r}
train <- read.csv("data/proteins_train.tsv", sep="\t", header=F)
test  <- read.csv("data/proteins_test.tsv",  sep="\t", header=F)
new   <- read.csv("data/proteins_new.tsv",   sep="\t", header=F)
```

```{r}
head(train)
```

## (b)

Estimate the vector of initial state probabilities $I$, the matrix of transition probabilities $T$ and the matrix for emission probabilities $E$ by maximum likelihood

## (c)

Estimate the stationary distribution $\pi$ of the Markov chain by solving the eigenvalue problem
(0.5 points) and by using a brute-force approach (0.5 points).

## (d)

Having estimated the parameters, i.e., the emission and transition matrices E, T and the
vector of initial state probabilities I, you can predict the latent state sequence Z of a protein’s
amino acid sequence X using the Viterbi algorithm. Use the Viterbi algorithm provided in
viterbi.r (carefully read the parameter description!) and iterate over each data.frame of
proteins test.tsv and proteins new.tsv row by row and use the amino acid sequence
to predict its secondary structure, which you add to the data.frame as a new column. Save
the extended data.frame of proteins new.tsv including the predicted secondary structure
as a tsv file and hand it in together with your pdf.

## (e)

Estimate confidence intervals for each parameter in I, E and T with bootstrapping. In a
single bootstrap run i estimate the probabilities for Ii
, Ei and Ti the same as before, but not
on the original data set proteins train.tsv, but on the resampled data set. I.e., sample
with replacement as many rows from proteins train.tsv as the original data set has. Run
a thousand bootstraps and compute the empirical 95% confidence intervals for each single
parameter in {Ii}i
, {Ei}i
and {Ti}i
.

## (f)

Use the following measure to compute the accuracy of the predicted secondary structure
P = (pi) for the data.frame of proteins test.tsv given the real secondary structure
S = (si):

$$
a(P, S) = 
\frac{1}{L} \sum_{i} \left\{\begin{aligned}
&1 & \text{if } p_i = s_i\\
&0 & \text{if } p_i \neq s_i
\end{aligned}\right.
$$

with sequence length L. Compute the accuracy for every protein in your data.frame and
store the accuracies in a vector. What is the accuracy of the Viterbi algorithm over all
sequences (i.e. call summary on the vector of accuracies).

## (g)

 Instead of using the Viterbi algorithm, now randomly guess secondary structures for all sequences. Compare the global accuracies of the Viterbi and the random approach and plot all
accuracy distributions using boxplots.