---
title: "Project 4 Membereship detection with profile HMMs"
date:  "`r format(Sys.time(), '%B %d, %Y')`"
author: "Team C - Minghang Li, Xiaocheng Yang, Xinyi Chen"
mainfont: "KpRoman"
editor_options: 
  chunk_output_type: console
  markdown: 
    wrap: 72
output:
  pdf_document:
    latex_engine: xelatex
    pandoc_args: "--pdf-engine-opt=-shell-escape"
    extra_dependencies: ["multirow", "float", "amsmath", "amsthm", "amssymb", "cancel", "unicode-math" ]
---

```{r dependencies, include=FALSE}
library(foreach)
library(doParallel)
```

```{r knitr_config, include=FALSE}
# the "pos" option is only available for neato or fdp
knitr::opts_chunk$set(engine.path = list(
  dot = Sys.which("neato") # path to neato
))
```

Problem 8-10 are all based on the MSA below:

```{r}

```

```{=latex}
\setlength{\tabcolsep}{10pt} % Default value: 6pt
\renewcommand{\arraystretch}{1.5}
\begin{table}[htbp]
\centering
\caption{Multiple sequence alignment for Problem 8-10}
\begin{tabular}{l|cccccc}
\multicolumn{1}{c|}{} & M & M & I       & I       & I       & M \\ \hline
bat                   & A & G & -       & -       & -       & C \\
rat                   & A & - & A       & G       & -       & C \\
cat                   & A & G & -       & -       & -       & C \\
gnat                  & - & G & A       & A       & A       & C \\
goat                  & A & G & -       & -       & A       & C \\ \hline
\multicolumn{1}{c|}{} & 1 & 2 & $\cdot$ & $\cdot$ & $\cdot$ & 3
\end{tabular}
\end{table}
```
# Problem 8: Estimating match emission probabilities

The count $E_i(a)$ and insert emission probability $e_i(a)$,
$\mathcal{a} \in \mathcal{A} = \{A, C, G, T\}$ are:

```{=latex}
\begin{table}[ht]
\centering
\caption{Estimated match emission probabilities of the profile HMM}
\begin{tabular}{ccccccc}
\hline
           & \multicolumn{2}{c}{\textbf{1}} & \multicolumn{2}{c}{\textbf{2}} & \multicolumn{2}{c}{\textbf{3}} \\ \hline
           & $E_1$          & $e_1$         & $E_2$          & $e_2$         & $E_3$          & $e_3$         \\ \hline
\textbf{A} & $5$            & $5/8$         & $1$            & $1/8$         & $1$            & $5/8$         \\
\textbf{C} & $1$            & $1/8$         & $1$            & $1/8$         & $6$            & $1/8$         \\
\textbf{G} & $1$            & $1/8$         & $5$            & $5/8$         & $1$            & $1/8$         \\
\textbf{T} & $1$            & $1/8$         & $1$            & $1/8$         & $1$            & $1/8$         \\ \hline
\end{tabular}
\end{table}
```
# Problem 9: Estimating insert emission probabilities

Since the contiguous insert states have the same position in the model,
the count $E_i(a)$ and insert emission probability $e_i(a)$,
$\mathcal{a} \in \mathcal{A} = \{A, C, G, T\}$ are:

```{=latex}
\begin{table}[ht]
\centering
\caption{Estimated insert emission probabilities of the profile HMM}
\begin{tabular}{ccccccc}
\hline
           & \multicolumn{2}{c}{\textbf{I}} \\ \hline
           & $E_1$          & $e_1$        \\ \hline
\textbf{A} & $6$            & $6/10$        \\
\textbf{C} & $1$            & $1/10$        \\
\textbf{G} & $2$            & $2/10$        \\
\textbf{T} & $1$            & $1/10$        \\ \hline
\end{tabular}
\end{table}
\pagebreak
```
# Problem 10: Estimating transition probabilities

The path of each sequence (\textcolor[HTML]{1f77b4}{bat},
\textcolor[HTML]{ff7f0e}{rat}, \textcolor[HTML]{2ca02c}{cat},
\textcolor[HTML]{d62728}{gnat}, \textcolor[HTML]{9467bd}{goat}) from
Begin to end is summarized below,

```{r dotHMM, engine='dot', fig.cap="HMM profile", echo=FALSE}
digraph HMM {
  rankdir=LR
  Begin [shape=square, color=blue, pos="0,0!"]
  m1 [shape=square, label=M, pos="1,0!"]
  m2 [shape=square, label=M, pos="2,0!"]
  m3 [shape=square, label=M, pos="3,0!"]
  End [shape=square, color=blue, pos="4,0!"]
  i2 [shape=diamond, label=I, pos="2,1!"]
  d1 [shape=circle, label=D, pos="1,2!"]
  d2 [shape=circle, label=D, pos="2,2!"]
 
  Begin -> m1 -> m2 -> m3 -> End [color="#1f77b4"] # bat
  Begin -> m1 -> d2 -> i2 -> i2 -> m3 -> End [color="#ff7f0e"] #rat
  Begin -> m1 -> m2 -> m3 -> End [color="#2ca02c"] # cat
  Begin -> d1 -> m2 -> i2 -> i2 -> i2 -> m3 -> End [color="#d62728"] #gnat
  Begin -> m1 -> m2 -> i2 -> m3 -> End [color="#9467bd"] #goat
}
```

Nodes that are not visited in any path have uniform probability of
transitting to its next states (as it only has pseudocounts). The
transition probabilities are summarized in the following table:

```{=latex}
\begin{table}[ht]
\centering
\caption{Estimated transition probabilities of the profile HMM}
\begin{tabular}{cccccccccc}
\hline
                  & \multicolumn{2}{c}{\textbf{0}} & \multicolumn{2}{c}{\textbf{1}} & \multicolumn{2}{c}{\textbf{2}} & \multicolumn{2}{c}{\textbf{3}} \\ \hline
                  & $E_0$          & $e_0$         & $E_1$          & $e_1$         & $E_2$          & $e_2$         & $E_3$          & $e_3$         \\ \hline
$M \rightarrow M$ & $5$            & $5/8$         & $4$            & $4/7$         & $3$            & $3/7$         & $6$            & $6/8$         \\
$M \rightarrow I$ & $1$            & $1/8$         & $1$            & $1/7$         & $3$            & $3/7$         & $1$            & $1/8$         \\
$M \rightarrow D$ & $2$            & $2/8$         & $2$            & $2/7$         & $1$            & $1/7$         & $1$            & $1/8$         \\ \hline
$I \rightarrow M$ & $1$            & $1/3$         & $1$            & $1/3$         & $4$            & $4/9$         & $1$            & $1/3$         \\
$I \rightarrow I$ & $1$            & $1/3$         & $1$            & $1/3$         & $4$            & $4/9$         & $1$            & $1/3$         \\
$I \rightarrow D$ & $1$            & $1/3$         & $1$            & $1/3$         & $1$            & $1/9$         & $1$            & $1/3$         \\ \hline
$D \rightarrow M$ & $1$            & $1/3$         & $2$            & $2/4$         & $1$            & $1/4$         & $1$            & $1/3$         \\
$D \rightarrow I$ & $1$            & $1/3$         & $1$            & $1/4$         & $2$            & $2/4$         & $1$            & $1/3$         \\
$D \rightarrow D$ & $1$            & $1/3$         & $1$            & $1/4$         & $1$            & $1/4$         & $1$            & $1/3$         \\ \hline
\end{tabular}
\end{table}
\pagebreak
```
# Problem 11: Protein family membership classification

## Import functions and read alignments

```{r load_source}
# import functions
source("code/profileHMM.R", local = knitr::knit_global())
```

```{r load_alignments}
# read alignments
GTPase <- parseAlignment("./data/GTP_binding_proteins.txt")
ATPase <- parseAlignment("./data/ATPases.txt")
```

## Learn HMM from two protein families

```{r lean_HMM}
HMM_GTPase <- learnHMM(GTPase)
HMM_ATPase <- learnHMM(ATPase)
```

## Identify position(s) with the highest match and insert emission frequencies over all symbols.

### In ATPase

The position(s) with the highest **match** emission frequency over all
symbols is/are:

```{r get_highest_match_emission_freq_ATPase}
mE_max_ATPase <- which(HMM_ATPase$mE == max(HMM_ATPase$mE, na.rm = TRUE), 
                       arr.ind = TRUE)
mE_max_ATPase[2] - 1
```

The position(s) with the highest **insert** emission frequency over all
symbols is/are:

```{r get_highest_insert_emission_freq_ATPase}
iE_max_ATPase <- which(HMM_ATPase$iE == max(HMM_ATPase$iE, na.rm = TRUE), 
                       arr.ind = TRUE)
iE_max_ATPase[2] - 1
```

```{r plot_highest_match_emission_freq_ATPase}
# generate barplot color sequence c() with length of alphabet
# all colors are gray except for position at mE_max_ATPase[1] is red
color_mE_max_ATPase <- c(rep("gray", length(HMM_ATPase$alphabet)))
color_mE_max_ATPase[mE_max_ATPase[1]] <- "red"
barplot(HMM_ATPase$mE[, mE_max_ATPase[2]],
        col = color_mE_max_ATPase,
        names = HMM_ATPase$alphabet,
        las = 2,
        main = paste("Match emission frequency distribution at position",
                     mE_max_ATPase[2] - 1),
        xlab = "Amino acids",
        ylab = "Match emission frequency")
```

```{r plot_highest_insert_emission_freq_ATPase}
# generate barplot color sequence c() with length of alphabet
# all colors are gray except for position at mE_max_ATPase[1] is red
color_iE_max_ATPase <- c(rep("gray", length(HMM_ATPase$alphabet)))
color_iE_max_ATPase[iE_max_ATPase[1]] <- "red"
barplot(HMM_ATPase$iE[, iE_max_ATPase[2]],
        col = color_iE_max_ATPase,
        names = HMM_ATPase$alphabet,
        las = 2,
        main = paste("Insertion emission frequency distribution at position",
                     iE_max_ATPase[2] - 1),
        xlab = "Amino acids",
        ylab = "Insertion emission frequency")
```

### In GTPase (GTP binding protein)

The position(s) with the highest **match** emission frequency over all
symbols is/are:

```{r get_highest_match_emission_freq_GTPase}
mE_max_GTPase <- which(HMM_GTPase$mE == max(HMM_GTPase$mE, na.rm = TRUE), 
                       arr.ind = TRUE)
mE_max_GTPase[2] - 1
```

The position(s) with the highest **insert** emission frequency over all
symbols is/are:

```{r get_highest_insert_emission_freq_GTPase}
iE_max_GTPase <- which(HMM_GTPase$iE == max(HMM_GTPase$iE, na.rm = TRUE), 
                       arr.ind = TRUE)
iE_max_GTPase[2] - 1
```

```{r plot_highest_match_emission_freq_GTPase}
# generate barplot color sequence c() with length of alphabet
# all colors are gray except for position at mE_max_ATPase[1] is red
color_mE_max_GTPase <- c(rep("gray", length(HMM_GTPase$alphabet)))
color_mE_max_GTPase[mE_max_GTPase[1]] <- "red"
barplot(HMM_GTPase$mE[, mE_max_GTPase[2]],
        col = color_mE_max_GTPase,
        names = HMM_GTPase$alphabet,
        las = 2,
        main = paste("Match emission frequency distribution at position",
                     mE_max_GTPase[2] - 1),
        xlab = "Amino acids",
        ylab = "Match emission frequency")
```

```{r plot_highest_insert_emission_freq_GTPase}
# generate barplot color sequence c() with length of alphabet
# all colors are gray except for position at mE_max_ATPase[1] is red
color_iE_max_GTPase <- c(rep("gray", length(HMM_GTPase$alphabet)))
color_iE_max_GTPase[iE_max_GTPase[1]] <- "red"
barplot(HMM_GTPase$iE[, iE_max_GTPase[2]],
        col = color_iE_max_GTPase,
        names = HMM_GTPase$alphabet,
        las = 2,
        main = paste("Insertion emission frequency distribution at position",
                     iE_max_GTPase[2] - 1),
        xlab = "Amino acids",
        ylab = "Insertion emission frequency")
```

## Classify unclassified proteins

```{r parse_proteins}
unclassified_proteins <- parseProteins(proteinsFile = "./data/Unclassified_proteins.txt")
```

```{r run_forward_algo}
num_cores <- detectCores()
registerDoParallel(num_cores)
log_odds_ATPase <- foreach(seq=unclassified_proteins, .combine = c) %dopar% {
  forward(HMM = HMM_ATPase, seq = seq)
}
log_odds_GTPase <- foreach(seq=unclassified_proteins, .combine = c) %dopar% {
  forward(HMM = HMM_GTPase, seq = seq)
}
stopImplicitCluster()
```

```{r plot_log_odds}
q <- log_odds_ATPase - log_odds_GTPase
names(q) <- NULL
plot(q, col=ifelse(q >= 0, "red", "blue"),
     xlab = "Sequence index",
     ylab = "Log odds ATPase vs GTPase")
legend("right", legend=c("ATPase", "GTPase"), col=c("red", "blue"), lty=1, cex=0.8)
```

We can see a clear separation between points that are classified as
ATPases and those classified as GTPases. The log ratio of ATPase vs
GTPase is always way larger than 0 or way smaller than 0, indicating
high likelihood.

The index of proteins that are classified as ATPases are:

```{r ATPase_index}
# remove list name
which(q >= 0)
```

The index of proteins that are classified as GTPases are:

```{r GTPase_index}
which(q < 0)
```
