---
title: "Project 11 NMF and multi-omics data integration"
date:  "`r format(Sys.time(), '%B %d, %Y')`"
author: "Team K - Minghang Li"
mainfont: "KpRoman"
monofont: "Fira Code Regular Nerd Font Complete"
output:
  pdf_document:
    latex_engine: xelatex
    extra_dependencies:  [ "amsthm", "amssymb", "cancel", "unicode-math" ]
---

```{r, include=FALSE}
library(NMF)
library(foreach)
library(doParallel)
```

# Problem 31: Uniqueness of NMF solutions

The solution is not necessarily unique. The easiest way is to give couter example.

Suppose we have:

$$
V = \begin{bmatrix}
  5 & 10
\end{bmatrix}
$$

Then it is possible to decompose it into given $k = 1$:

$$
H = \begin{bmatrix}
  1 & 2
\end{bmatrix}, \quad
W = \begin{bmatrix}
  5
\end{bmatrix}
$$

However, suppose we have $\tilde{W} = \begin{bmatrix} 2 \end{bmatrix}$, we can also find $\tilde{H}$ such that:

$$
\tilde {H} = \tilde{W}^{-1} W H = \begin{bmatrix}
  5/2 & 5
\end{bmatrix}
$$

So that $WH = \tilde{W}\tilde{H}$ but $W \neq \tilde{W}$ and $H \neq \tilde{H}$. In general if we have square $\tilde{W}$ and it is inversible the solution is not unique.

(Of course I'm not satisfied with this proof because it only proves the non-uniqueness when $K = M$. I'm not sure how to tackle this problem in a more general scenario though...)

# Problem 32: NMF of spatial gene expression patterns

## Subproblem 1

```{r}
# load DrosophilaExpressions.rda
load("DrosophilaExpressions.rda")
```

Display the first 16 observations as elliptical images.

```{r}
imageBatchDisplay(V[, 1:16])
```

### Factorization using `rank = 15`

Compute a factorization $V \approx \hat{V} = WH$ with `rank=15, seed=123` and default `method="brunet"`.

```{r}
res <- nmf(V, rank = 15, method = "brunet", seed = 123)
```

Report the KL divergence of the approximated data matrix $\hat{V}$.

```{r}
summary(res)
```

Display the first 16 columns of $\hat{V}$ as imagese.

```{r}
V.hat <- fitted(res)
imageBatchDisplay(V.hat[, 1:16])
```

Display all computed basis patterns (columns of W).

```{r, out.width="80%"}
# use basismap (from the documentation)
basismap(res, subsetRow=TRUE)
```

### Factorization using `rank = 10`

```{r}
res.10 <- nmf(V, rank = 10, method = "brunet", seed = 123)
# report KL divergence (note the "residuals")
summary(res.10)
```

```{r}
# visualize V hat
V.hat.10 <- fitted(res)
imageBatchDisplay(V.hat.10[, 1:16])
```

```{r, out.width="80%"}
# display W
basismap(res.10, subsetRow=TRUE)
```

## Subproblem 2


```{r}
W.10 <- basis(res.10)
H.10 <- coef(res.10)
```

### Gene `Mkp3`

```{r}
imageBatchDisplay(
  W.10,
  imgNames = H.10[, "Mkp3"],
  font = 1
)
```

### Gene `CG31909`

```{r}
imageBatchDisplay(
  W.10,
  imgNames = H.10[, "CG31909"],
  font = 1
)
```

# Problem 33: Implementing NMF from scratch

```{r}
NMF <- function(V, k, 
                W = NULL, H = NULL, max_iter = 10000,
                seed = 42) {
  set.seed(seed)
  M <- dim(V)[[1]]
  N <- dim(V)[[2]]
  if (is.null(W)) {
    W <- matrix(runif(M * k), nrow = M, ncol = k)
  }
  
  if (is.null(H)) {
    H <- matrix(runif(k * N), nrow = k, ncol = N)
  }
  
  W_new <- W
  H_new <- H
  to_log <- V /(W %*% H)
  to_log <- ifelse(to_log != 0, log(to_log), 0)
  kl_div <- sum(V * to_log - V + W %*% H)
  prev_kl_div <- kl_div
  all_kl <- foreach (iter=1:max_iter, .combine = c) %do% {
    H_new <- H * (t(W) %*% (V/(W %*% H))) / (colSums(W))
    W_new <- W * ((V/(W %*% H_new)) %*% t(H_new)) / (rowSums(H_new))
    
    to_log <- V /(W_new %*% H_new)
    to_log <- ifelse(to_log != 0, log(to_log), 0)
    kl_div <- sum(V * to_log - V + W_new %*% H_new)

    prev_kl_div <- kl_div
    W <- W_new
    H <- H_new
    kl_div
  }
  
  return(list(V.hat=W_new %*% H_new, W=W_new, H=H_new, KLdiv=kl_div, allKL=all_kl))
}
```

```{r}
res <- NMF(V, 10, max_iter=1000)
```

```{r}
res$KLdiv
```

```{r}
imageBatchDisplay(res$V.hat[, 1:16])
```