---
title: "Project 8 Nested Effect Models"
date:  "`r format(Sys.time(), '%B %d, %Y')`"
author: "Team K - Minghang Li"
mainfont: "KpRoman"
monofont: "Fira Code Regular Nerd Font Complete"
output:
  pdf_document:
    latex_engine: xelatex
    extra_dependencies:  [ "amsthm", "amssymb", "cancel", "unicode-math" ]
---

```{r, include=FALSE}
library(mnem)
```

# Problem 20: Classical NEMs

## Subproblem 1

### Construct transitive closure and define $\Phi$

Define a function to make any $\Phi$ matrix transitive closed by powering it up until convergence.

```{r}
transitive_closify <- function(phi) {
  old_phi <- phi
  while (TRUE) {
    new_phi <- old_phi %*% phi
    new_phi[new_phi > 0] <- 1
    if (isTRUE(all.equal(new_phi, old_phi))) {
      break
    }
    old_phi <- new_phi
  }
  return(new_phi)
}
```

Construct $\Phi$ for Model (a).

```{r}
phi_a <- array(
  dim = c(5, 5),
  dimnames = list(c("S1", "S2", "S3", "S4", "S5"),
                  c("S1", "S2", "S3", "S4", "S5"))
)
#                  S1, S2, S3, S4, S5
phi_a["S1", ] <- c( 1,  0,  1,  1,  0)
phi_a["S2", ] <- c( 0,  1,  0,  0,  1)
phi_a["S3", ] <- c( 0,  0,  1,  1,  1)
phi_a["S4", ] <- c( 0,  0,  0,  1,  1)
phi_a["S5", ] <- c( 0,  0,  0,  0,  1)
```

```{r}
phi_a <- transitive_closify(phi_a)
phi_a
```

Construct $\Phi$ for Model (b).

```{r}
phi_b <- array(
  dim = c(5, 5),
  dimnames = list(c("S1", "S2", "S3", "S4", "S5"),
                  c("S1", "S2", "S3", "S4", "S5"))
)
#                  S1, S2, S3, S4, S5
phi_b["S1", ] <- c( 1,  0,  0,  1,  0)
phi_b["S2", ] <- c( 0,  1,  0,  0,  1)
phi_b["S3", ] <- c( 1,  0,  1,  1,  1)
phi_b["S4", ] <- c( 0,  0,  0,  1,  1)
phi_b["S5", ] <- c( 0,  0,  0,  0,  1)
```

```{r}
phi_b <- transitive_closify(phi_b)
phi_b
```

### Define $\Theta$

Define $\Theta$ for Model (a).

```{r}
theta_a <- array(
  dim = c(5, 6),
  dimnames = list(c("S1", "S2", "S3", "S4", "S5"),
                  c("E1", "E2", "E3", "E4", "E5", "E6"))
)
#                    E1, E2, E3, E4, E5, E6
theta_a["S1", ] <- c( 0,  0,  0,  0,  0,  0)
theta_a["S2", ] <- c( 0,  0,  0,  1,  0,  1)
theta_a["S3", ] <- c( 1,  1,  0,  0,  0,  0)
theta_a["S4", ] <- c( 0,  0,  1,  0,  0,  0)
theta_a["S5", ] <- c( 0,  0,  0,  0,  1,  0)
theta_a
```

Define $\Theta$ for Model (b).

```{r}
theta_b <- array(
  dim = c(5, 6),
  dimnames = list(c("S1", "S2", "S3", "S4", "S5"),
                  c("E1", "E2", "E3", "E4", "E5", "E6"))
)
#                    E1, E2, E3, E4, E5, E6
theta_b["S1", ] <- c( 1,  1,  0,  0,  0,  0)
theta_b["S2", ] <- c( 0,  0,  0,  1,  0,  1)
theta_b["S3", ] <- c( 0,  0,  0,  0,  0,  0)
theta_b["S4", ] <- c( 0,  0,  1,  0,  0,  0)
theta_b["S5", ] <- c( 0,  0,  0,  0,  1,  0)
theta_b
```

### Determine the corresponding expected effect patterns ($F$)

```{r}
F_a <- phi_a %*% theta_a
F_a
```
```{r}
F_b <- phi_b %*% theta_b
F_b
```

## Subproblem 2

If we assume no noise (no false positives and false negatives)... then the $D$ matrix is simply the $F$ matrix transpose.

```{r}
D_a <- t(F_a)
D_a
```

```{r}
D_b <- t(F_b)
D_b
```

Given the discrete data $D_a$ and $D_b$ (sorry for the different notation from the exercise pdf) it's not possible to tell apart the two models because they are identical.

```{r}
all.equal(D_a, D_b)
```

## Subproblem 3

Calculate the marginal log-likelihood ratio (network score) given the data by setting the false positive rate to be 5\% and the false negative rate to be 1\%.

```{r}
network_score_a <- scoreAdj(D_a, 
                            adj = phi_a, 
                            method = "disc",
                            fpfn = c(0.05, 0.01)
                            )$score
network_score_a
```

```{r}
network_score_b <- scoreAdj(D_b, 
                            adj = phi_b, 
                            method = "disc",
                            fpfn = c(0.05, 0.01)
                            )$score
network_score_b
```

# Problem 21: Hidden Markov NEMs

## Subproblem 1

## Subproblem 2

# Problem 22: Mixture NEMs

## Subproblem 1

## Subproblem 2

### (a) Compute expected effect pattern $(\rho^T\phi_k\theta_k)^T$

### (b) 

## Subproblem 3






